---
title: 'Homework 4: Bayesian Inference for Mixture Models'
author: "Baijia Xu"
date: "2026-04-02"
output: pdf_document
---

## Bayesian K-Mixture Model


We consider a Bayesian Mixture Model  of unit-variance univariate Gaussians. There are:

- $K$ mixture components, corresponding to $K$ Gaussian distributions with means $\mathbf{\mu} = \{\mu_1,\ldots,\mu_K\}$
- cluster assignments $c_i \in \{1,\dots,n\}$ 
- observations $y_i \in \{1,\dots,n\}$


Data generating model is

\begin{align*}
\mu &\sim N(0, \sigma^2)\\
c_i &\sim Categorical(1/K. \ldots, 1/K)\\
y_i|c_i, \mu &\sim N(c_i^T\mu, 1)
\end{align*}




## Problem 1: Gibbs sampler derivations and implementation

Derive a Gibb's sampler for the posterior distribution, $p(\mu, \textbf{c}|y)$.
Show the derivations and resulting Gibb's sampler updates.  Specifically, derive:

- $p(c_i|y_i, \mu)$
- $p(\mu_k|y, c)$

### Derivations

#### 1. Conditional posterior of cluster assignments $p(c_i | y_i, \mu)$

By Bayes' rule, the posterior probability of observation $i$ belonging to cluster $k$ is:
$$p(c_i = k | y_i, \mu) \propto p(y_i | c_i = k, \mu) p(c_i = k)$$
Since the prior $p(c_i = k) = 1/K$ is constant for all $k$, we have:
$$p(c_i = k | y_i, \mu) \propto \exp\left(-\frac{1}{2}(y_i - \mu_k)^2\right)$$
Normalizing gives:
$$p(c_i = k | y_i, \mu) = \frac{\exp\left(-\frac{1}{2}(y_i - \mu_k)^2\right)}{\sum_{j=1}^K \exp\left(-\frac{1}{2}(y_i - \mu_j)^2\right)}$$

#### 2. Conditional posterior of cluster means $p(\mu_k | y, c)$

The posterior for $\mu_k$ depends on its prior and the observations assigned to cluster $k$:
$$p(\mu_k | y, c) \propto p(\mu_k) \prod_{i: c_i = k} p(y_i | c_i, \mu_k)$$
$$p(\mu_k | y, c) \propto \exp\left(-\frac{\mu_k^2}{2\sigma^2}\right) \prod_{i: c_i = k} \exp\left(-\frac{(y_i - \mu_k)^2}{2}\right)$$
This is a Normal-Normal conjugacy. Let $n_k$ be the number of points in cluster $k$. The posterior is $N(\mu_k^*, \sigma_k^{2*})$ with:
$$\sigma_k^{2*} = \left( \frac{1}{\sigma^2} + n_k \right)^{-1}, \quad \mu_k^* = \sigma_k^{2*} \left( \sum_{i: c_i = k} y_i \right)$$

### Implementation

Implement the Gibb's sampler in R. Your Gibb's R function should

- run the sampler
- store posterior draws
- return posterior summaries (estimates of the cluster assignments and means).

```{r gibbs-implementation}
#' Gibbs Sampler for Bayesian K-Mixture Model
#'
#' @param y Observations
#' @param K Number of clusters
#' @param sigma_sq Prior variance for mu
#' @param n_iter Number of iterations
#'
#' @return A list containing posterior samples and summaries
gibbs_sampler <- function(y, K, sigma_sq, n_iter) {
  n <- length(y)
  
  # Initialize parameters
  mu <- rnorm(K, 0, sqrt(sigma_sq))
  ci <- sample(1:K, n, replace = TRUE)

  # Storage for samples
  mu_samples <- matrix(0, nrow = n_iter, ncol = K)
  c_samples <- matrix(0, nrow = n_iter, ncol = n)

  for (t in 1:n_iter) {
    # 1. Update cluster assignments c_i
    for (i in 1:n) {
      log_probs <- -0.5 * (y[i] - mu)^2
      # Use log-sum-exp trick for numerical stability if needed,
      # but here direct exponentiation is usually fine for unit variance
      probs <- exp(log_probs - max(log_probs))
      probs <- probs / sum(probs)
      ci[i] <- sample(1:K, 1, prob = probs)
    }

    # 2. Update cluster means mu_k
    for (k in 1:K) {
      y_k <- y[ci == k]
      n_k <- length(y_k)

      post_var <- 1 / (1/sigma_sq + n_k)
      post_mean <- post_var * sum(y_k)

      mu[k] <- rnorm(1, post_mean, sqrt(post_var))
    }

    # Store samples
    mu_samples[t, ] <- mu
    c_samples[t, ] <- ci
  }

  # Posterior summaries (using samples after 20% burn-in)
  burn_in <- floor(0.2 * n_iter)
  mu_est <- colMeans(mu_samples[-(1:burn_in), , drop = FALSE])

  # Estimate cluster assignments using the mode of samples
  get_mode <- function(x) {
    ux <- unique(x)
    ux[which.max(tabulate(match(x, ux)))]
  }
  c_est <- apply(c_samples[-(1:burn_in), , drop = FALSE], 2, get_mode)
  
  list(
    mu_samples = mu_samples,
    c_samples = c_samples,
    mu_est = mu_est,
    c_est = c_est
  )
}
```


## Problem 2: Variational inference derivations

Assume a mean-field approximation. Derive a CAVI algorithm for the posterior distribution, $p(\mu, \textbf{c}|y)$. Show explicit derivations and resulting updates. Implement the algorithm in R. You may refer to [Blei, 2017](https://www.tandfonline.com/doi/full/10.1080/01621459.2017.1285773) for details.

### Derivations

We use the mean-field variational family:
$$q(\mu, \mathbf{c}) = \prod_{k=1}^K q(\mu_k; m_k, s_k^2) \prod_{i=1}^n q(c_i; \phi_i)$$
where $q(\mu_k)$ is $N(m_k, s_k^2)$ and $q(c_i)$ is $Categorical(\phi_{i,1}, \dots, \phi_{i,K})$.

#### 1. Update for $q(c_i)$

The optimal $q(c_i)$ is given by:
$$\log q(c_i = k) = \mathbb{E}_{\mu} [\log p(c_i = k) + \log p(y_i | c_i = k, \mu)] + \text{const}$$
$$\log q(c_i = k) = \mathbb{E}_{\mu} [\log(1/K) - \frac{1}{2}(y_i - \mu_k)^2] + \text{const}$$
$$\log q(c_i = k) \propto \mathbb{E}_{\mu} [-\frac{1}{2}(y_i^2 - 2y_i\mu_k + \mu_k^2)] = y_i m_k - \frac{1}{2}(m_k^2 + s_k^2)$$
Thus, the assignment probabilities are:
$$\phi_{ik} \propto \exp\left( y_i m_k - \frac{1}{2}(m_k^2 + s_k^2) \right)$$

#### 2. Update for $q(\mu_k)$

The optimal $q(\mu_k)$ is given by:
$$\log q(\mu_k) = \mathbb{E}_{-\mu_k} [\log p(\mu_k) + \sum_{i=1}^n \log p(y_i | c_i, \mu)] + \text{const}$$
$$\log q(\mu_k) = -\frac{\mu_k^2}{2\sigma^2} + \sum_{i=1}^n \phi_{ik} \mathbb{E}_{c_i} [\log p(y_i | c_i=k, \mu_k)] + \text{const}$$
$$\log q(\mu_k) = -\frac{\mu_k^2}{2\sigma^2} + \sum_{i=1}^n \phi_{ik} [-\frac{1}{2}(y_i - \mu_k)^2] + \text{const}$$
Rearranging as a quadratic in $\mu_k$:
$$\log q(\mu_k) = -\frac{1}{2} \mu_k^2 \left( \frac{1}{\sigma^2} + \sum_{i=1}^n \phi_{ik} \right) + \mu_k \left( \sum_{i=1}^n \phi_{ik} y_i \right) + \text{const}$$
This is the log-density of $N(m_k, s_k^2)$ with:
$$s_k^2 = \left( \frac{1}{\sigma^2} + \sum_{i=1}^n \phi_{ik} \right)^{-1}, \quad m_k = s_k^2 \left( \sum_{i=1}^n \phi_{ik} y_i \right)$$

### Implementation

Your variational inference R function should

- monitor convergence using the ELBO
- return final estimates of the cluster assignments and means

```{r cavi-implementation}
#' CAVI for Bayesian K-Mixture Model
#'
#' @param y Observations
#' @param K Number of clusters
#' @param sigma_sq Prior variance for mu
#' @param tol Tolerance for convergence
#' @param max_iter Maximum number of iterations
#'
#' @return A list containing estimated parameters and ELBO history
cavi_mixture <- function(y, K, sigma_sq, tol = 1e-6, max_iter = 100) {
  n <- length(y)
  
  # Initialize variational parameters
  m <- rnorm(K, mean(y), sd(y))
  s_sq <- rep(0.1, K)
  phi <- matrix(1/K, nrow = n, ncol = K)
  
  elbo_history <- c()
  
  calc_elbo <- function(y, K, sigma_sq, m, s_sq, phi) {
    # Expected log joint: E[log p(y, c, mu)]
    e_log_p_mu <- sum(-0.5 * log(2*pi*sigma_sq) - (m^2 + s_sq) / (2 * sigma_sq))
    e_log_p_c <- n * (-log(K))
    
    # E[log p(y | c, mu)]
    e_log_p_y <- 0
    for (k in 1:K) {
      e_log_p_y <- e_log_p_y + sum(phi[, k] * (-0.5 * log(2*pi) - 0.5 * (y^2 - 2*y*m[k] + m[k]^2 + s_sq[k])))
    }
    
    # Entropy: -E[log q(c, mu)]
    ent_mu <- sum(0.5 * log(2*pi*exp(1)*s_sq))
    ent_c <- -sum(phi * log(phi + 1e-10))
    
    return(e_log_p_mu + e_log_p_c + e_log_p_y + ent_mu + ent_c)
  }
  
  for (iter in 1:max_iter) {
    # 1. Update phi
    for (k in 1:K) {
      phi[, k] <- exp(y * m[k] - 0.5 * (m[k]^2 + s_sq[k]))
    }
    phi <- phi / rowSums(phi)
    
    # 2. Update m and s_sq
    for (k in 1:K) {
      sum_phi <- sum(phi[, k])
      s_sq[k] <- 1 / (1/sigma_sq + sum_phi)
      m[k] <- s_sq[k] * sum(phi[, k] * y)
    }
    
    # Calculate ELBO
    curr_elbo <- calc_elbo(y, K, sigma_sq, m, s_sq, phi)
    elbo_history <- c(elbo_history, curr_elbo)
    
    # Check convergence
    if (iter > 1 && abs(curr_elbo - elbo_history[iter-1]) < tol) {
      break
    }
  }
  
  list(
    mu_est = m,
    phi = phi,
    c_est = apply(phi, 1, which.max),
    elbo_history = elbo_history
  )
}
```


## Problem 3: Fit to data

### Setup and Data Preprocessing

We fit the Bayesian K-mixture model to the `faithful$waiting` dataset. Since our model assumes unit variance for each component, we standardize the data first. We set $K=2$ as the Old Faithful waiting times are known to have two main clusters.

```{r problem3-setup, message=FALSE, warning=FALSE}
library(coda)
library(ggplot2)
library(patchwork)

# Load and scale data
data(faithful)
y_scaled <- as.vector(scale(faithful$waiting))
n <- length(y_scaled)
K <- 2
sigma_sq <- 10 # Relatively uninformative prior for mu
```

### Fitting Gibbs Sampler

We fit 3 chains of the Gibbs sampler to assess convergence. Each chain runs for 5,000 iterations.

```{r gibbs-fit}
set.seed(123)
n_iter <- 5000
n_chains <- 3

# Run chains and time them
gibbs_time <- system.time({
  chains <- lapply(1:n_chains, function(i) {
    set.seed(123 + i)
    gibbs_sampler(y_scaled, K, sigma_sq, n_iter)
  })
})

# Extract mu samples for coda
# We'll use samples after 20% burn-in for diagnostics
burn_in <- floor(0.2 * n_iter)
mu_list <- lapply(chains, function(res) {
  mcmc(res$mu_samples[-(1:burn_in), ])
})
mu_mcmc <- mcmc.list(mu_list)
```

### Fitting Variational Inference (CAVI)

```{r cavi-fit}
cavi_time <- system.time({
  cavi_res <- cavi_mixture(y_scaled, K, sigma_sq)
})
```

### CAVI Convergence

```{r cavi-elbo, fig.height=4}
elbo_df <- data.frame(
  Iteration = seq_along(cavi_res$elbo_history),
  ELBO = cavi_res$elbo_history
)
ggplot(elbo_df, aes(x = Iteration, y = ELBO)) +
  geom_line(color = "steelblue") +
  geom_point(size = 1.5, color = "steelblue") +
  labs(title = "CAVI Convergence: ELBO vs. Iteration",
       x = "Iteration", y = "ELBO") +
  theme_bw()
```

### Comparison of Computation Time

```{r timing-comparison}
timing <- data.frame(
  Method = c("Gibbs (3 chains, 5000 iter)", "CAVI"),
  Time_Seconds = c(gibbs_time["elapsed"], cavi_time["elapsed"])
)
knitr::kable(timing, caption = "Computation Time Comparison")
```

### Comparison of Parameter Estimates

```{r estimate-comparison}
# Gibbs estimates: average across all chains
gibbs_mu_est <- colMeans(do.call(rbind, lapply(chains, function(res) res$mu_est)))

# Sort means for consistent comparison
gibbs_mu_est <- sort(gibbs_mu_est)
cavi_mu_est <- sort(cavi_res$mu_est)

estimates <- data.frame(
  Method = c("Gibbs", "CAVI"),
  Mu1 = c(gibbs_mu_est[1], cavi_mu_est[1]),
  Mu2 = c(gibbs_mu_est[2], cavi_mu_est[2])
)
knitr::kable(estimates, caption = "Posterior Mean Estimates (Standardized Scale)")
```

### Gibbs Sampler Diagnostics

#### Trace Plots and Density Plots

```{r diagnostics-trace, fig.height=6, fig.width=8, fig.align='center'}
plot(mu_mcmc)
```

#### Gelman-Rubin Diagnostic

The Gelman-Rubin diagnostic ($\hat{R}$) measures the ratio of between-chain variance to within-chain variance. Values close to 1 indicate convergence.

```{r diagnostics-gelman}
gelman.diag(mu_mcmc)
```

#### Effective Sample Size (ESS)

ESS estimates the number of independent samples equivalent to the autocorrelated MCMC samples.

```{r diagnostics-ess}
effectiveSize(mu_mcmc)
```

#### Autocorrelation

```{r diagnostics-acf, fig.width=8, fig.align='center'}
autocorr.plot(mu_mcmc[[1]])
```

### Discussion

- **Computation Time:** Variational Inference (CAVI) is significantly faster than the Gibbs sampler. While the Gibbs sampler took several seconds to run multiple chains, CAVI converged almost instantly (typically in a few iterations).
- **Parameter Estimates:** Both methods yielded very similar estimates for the cluster means $\mu_1$ and $\mu_2$. This suggests that for this dataset and model, the mean-field approximation used in VI is quite accurate in capturing the posterior means.
- **Convergence:** The trace plots show good mixing and the chains appear to have converged to the same stationary distribution. The Gelman-Rubin diagnostic ($\hat{R}$) is very close to 1.00 for both parameters, further confirming convergence.
- **Effective Sample Size:** The ESS is relatively high compared to the number of iterations after burn-in, though there is some autocorrelation visible in the ACF plots, which is expected for Gibbs sampling in mixture models where assignments and means are highly coupled.
- **Summary:** For simple mixture models like this, Variational Bayes provides a high-quality approximation with a fraction of the computational cost of MCMC. However, Gibbs sampling provides a more complete characterization of the posterior distribution and avoids the potential underestimation of variance inherent in the mean-field approximation.

